---
title: "Prepare HSLS Data for Analysis"
author: "Kristian Edosomwan"
date: "2026-04-17"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Data Engineering
## Load Libraries

```{r libraries}
library(haven)
library(dplyr)
library(labelled)
library(psych)
library(broom)
```

# Load Dataset
```{r}
hsls_raw <- read_dta("hsls_17_student_pets_sr_v1_0.dta")
```

# Subset Dataset Based on Grant Application and LowerCase
```{r}
hsls_subset <- hsls_raw %>% dplyr::select(
# Student Demographics & Background
STU_ID, X1SES, X1SESQ5, X1RACE, X1SEX, X1SES_U, X2SES, X2SES_U, X2RACE, X2SESQ5, X2SEX, X2SESQ5_U, X1PAREDU, X1POVERTY185, X2PAREDU, X2POVERTY185, X1POVERTY130, X1POVERTY, X1FAMINCOME, X2FAMINCOME, X1PAR1RACE, X1PAR2RACE, P1ELLEVER, X1DUALLANG,
# Student Mathematics Scores and STEM GPA
S1M8GRADE, S1S8GRADE, X1TXMQUINT, X1TXMTH, X1TXMTSCOR, X2TXMQUINT, X2TXMTH, X2TXMTSCOR, X3TGPAHIMTH, X3TGPAHISCI, X3TGPAMAT, X3TGPASTEM,
# Student Mathematics Scales [in Red] and Component Items
X1MTHEFF, S1MTESTS, S1MTEXTBOOK, S1MSKILLS, S1MASSEXCL,
X1MTHID, S1MPERSON1, S1MPERSON2,
X1MTHINT, S1MENJOYING, S1MWASTE, S1MBORING, S1FAVSUBJ, S1LEASTSUBJ, S1MENJOYS,
X1MTHUTI, S1MUSELIFE, S1MUSECLG, S1MUSEJOB,
X2MTHEFF, S2MTESTS, S2MTEXTBOOK, S2MSKILLS, S2MASSEXCL,
X2MTHID, S2MPERSON1, S2MPERSON2,
X2MTHINT_R, S2MENJOYING, S2MWASTE, S2MBORING, S2FAVSUBJ, S2MENJOYS,
X2MTHUTI, S2MUSELIFE, S2MUSECLG, S2MUSEJOB,
X2MEFFORT, S2MATTENTION, S2MONTIME, S2MSTOPTRYING, S2MGETBY,
# Student Perceptions of Mathematics Teacher
S1MTCHRESPCT, S1MTCHFAIR, S1MTCHCONF, S1MTCHMISTKE, S1MTCHTREAT, S1MTCHINTRST, S1MTCHMFDIFF, S1MTCHEASY, S2MTCHTREAT, S2MTCHINTRST, S2MTCHEASY, S2MTCHTHINK, S2MTCHGIVEUP,
# Mathematics and Science Secondary Coursework
S1M8, S1S8, X3T1CREDALG1, X3T1CREDALG2, X3T1CREDCALC, X3T1CREDCHEM, X3T1CREDPHYS, X3T1CREDPREC, X3T1CREDSTAT, X3TCREDAPMTH, X3TCREDAPSCI, X3TCREDMAT, X3TCREDMTSC, X3TCREDSTEM, X3THIMATH, X3THIMATH9, X3THISCI, X3THISCI9, X3TWHENALG1, S2HIMATH12, S2HISCIENCE12,
# Postsecondary STEM major and coursework
X4RFDGMJSTEM, X5STEMCRED, X5STEM1ATT, X5STEM1ERN, X5STEM1GPA, X5STEM3ATT, X5STEM3ERN, X5STEM3GPA,
# Parent Expectations & Educational Engagement
P1EDUASPIRE, P1EDUEXPECT, P1FUNDRAISE, P1HONORS, P1MTHHWEFF, P1PTCONFER, P1PTOMTG, P1COUNSELOR, P1SCHEVENT, P1SCHMTG, P1SCIFAIR, P1SCIHWEFF, P1SCIPROJ, P1STEMDISC, P1VOLUNTEER, P1ABLEBA, P2EDUASP, P2EDUEXP, P2ABLEBA, P2DISCCOURSES, P2MTHHWEFF, P2SCIHWEFF, P2SCIPROJ, P2STEMDISC, S1DADTALKM, S1DADTALKS, S1MOMTALKM, S1MOMTALKS,
# Math Teacher Expectations and Beliefs [Scales in Red]
X1TMCOMM, M1SHRIDEAS, M1WORKSHOP, M1SHRSTWRK, M1SHRLESSONS, M1SHRBELIEFS, M1SHRMTHDS, M1SHRELL, M1SHRAPPRCH, M1SHRCONTENT, M1EFFECTIVE, M1MENTOR, M1CHAIR, 
X1MTHEFF, M1FAMILY, M1DISCIPLINE, M1STUACHIEVE, M1PARENT, M1RETAIN, M1REDIRECT, M1GETTHRU, M1HOMEFX, 
X1TMEXP, M1TEACHING, M1LEARNING, M1BELIEVE, M1CLEARGOALS, M1GIVEUP, M1CARE, M1EXPECT, M1WORKHARD, 
X1TMPRINC, M1PRESSURES, M1POORJOBRES, M1PSETSPRIO, M1PSCHVISION, M1PCOMEXP, M1PINNOVATE, M1PCONSULTS, 
X1TMRESP, M1TSCHDISC, M1TIMPROVE, M1TSETSTDS, M1TSELFDEV, M1THELPBEST, M1TALLLEARN, M1TFAIL, 
X1TSCOMM, M1SHRIDEAS, M1WORKSHOP, M1SHRSTWRK, M1SHRLESSONS, M1SHRBELIEFS, M1SHRMTHDS, M1SHRELL, M1SHRAPPRCH, M1SHRCONTENT, M1EFFECTIVE, M1MENTOR, M1CHAIR,
# Math Teacher Instructional Practices/Focus
M1GROUP, M1ASSIGN, M1INTEREST, M1CONCEPTS, M1ALGORITHM, M1COMPSKILLS, M1PROBLEM, M1REASON, M1IDEAS, M1PREPARE, M1LOGIC, M1HISTORY, M1EXPLAIN, M1BUSINESS, M1COMPUTE, M1TEST, M1AUTONOMY, 
# School Characteristics 
X1CONTROL, X2CONTROL, X3CONTROL, X4CONTROL, X1LOCALE, X2LOCALE, X3LOCALE, X4LOCALE, X4PS1CTRL, 
# Sampling Weights
W1STUDENT, W1PARENT, W1MATHTCH, W2STUDENT, W2W1STU, W2PARENT, W2W1PAR, W3STUDENT, W3W1STU, W3W1W2STU, W3W2STU, W3HSTRANS, W3STUDENTTR, W3W1STUTR, W3W1W2STUTR, W3W2STUTR, W4STUDENT, W4W1STU, W4W1STUP1, W4W1STUP1P2, W4W1W2W3STU, W5PSTRANS, W5PSRECORDS, W5W1W2W3W4PSTRANS, W5W1W2W3W4PSRECORDS, W3W1MATHTCH
)

names(hsls_subset) <- tolower(names(hsls_subset))
```

## View Data Original Data Subset

```{r}
library(sjPlot)

# drop nonresponses based on having unit non-response for highest math class in 8th grade, which we are using as our first time point, and highest math class in spring 2012, and higest math credit
hsls <- hsls_subset %>% dplyr::filter(s1m8 != -8, x3thimath9 !=-8, s2himath12!=-8, x3thimath!=-8)

# view data before replacing missing values
## view_df(hsls, show.frq = TRUE, show.prc = TRUE, sort = TRUE)

```

## View Data Subset with Missing Values
```{r}
# replace missing values
hsls[hsls == -9] <- NA # missing
hsls[hsls == -8] <- NA # non-response
hsls[hsls == -7] <- NA # legit skip
hsls[hsls == -6] <- NA # not applicable
hsls[hsls == -4] <- NA # not administered
hsls[hsls == -3] <- NA # carry through missing
hsls[hsls == -1] <- NA # don't know

# drop unused labels
hsls <- labelled::drop_unused_value_labels(hsls)

# view data after replacing missing values
view_df(hsls, show.frq = TRUE, show.prc = TRUE, show.na = TRUE,  sort = TRUE)

```
# Variable Manipulation
```{r}
# ---------------------------
# Outcome variable
# ---------------------------
# par_says_honors: binary numeric (0/1)
hsls <- hsls %>%
  mutate(
    par_says_honors = case_when(
      p1honors == 1 ~ "1 yes",
      p1honors == 0 ~ "0 no",
      TRUE ~ NA
    )
  )

# ---------------------------
# Treatment variables
# ---------------------------
hsls <- hsls %>%
  mutate(
    # adv_math: binary numeric (0 = not advanced, 1 = advanced)
    adv_math = case_when(
      x3thimath9 <= 4 ~ "1 yes",
      x3thimath9 >= 5 ~ "0 no"
    ),

    # math_3level: ordinal categorical (low → high)
    math_3level = ordered(
      case_when(
        x3thimath9 <= 3 ~ "1 remedial",
        x3thimath9 == 4 ~ "2 on-level",
        x3thimath9 >= 5 ~ "3 advanced"
      ),
      levels = c("1 remedial", "2 on-level", "3 advanced")
    ),

    # math_4level: ordinal categorical (low → high)
    math_4level = ordered(
      case_when(
        x3thimath9 <= 3 ~ "1 remedial",
        x3thimath9 == 4 ~ "2 on-level",
        x3thimath9 == 5 ~ "3 advanced",
        x3thimath9 >= 6 ~ "4 double advanced"
      ),
      levels = c("1 remedial", "2 on-level", 
                 "3 advanced", "4 double advanced")
    )
  )

# ---------------------------
# Controls
# ---------------------------
hsls <- hsls %>%
  mutate(
    # urban: binary numeric
    urban = ifelse(x1locale == 1, "1 urban", "0 non-urban"),

    # math_score: continuous numeric
    math_score = ifelse(x1txmquint < 0, NA, x1txmtscor),

    # public: binary numeric (1 = public, 0 = private)
    public = case_when(
      x1control == 1 ~ "1 public",
      x1control == 2 ~ "0 private"
    )
  )

# ---------------------------
# Demographics
# ---------------------------
hsls <- hsls %>%
  mutate(
    # first_gen: binary numeric
    first_gen = ifelse(x1paredu <= 3, 1,0),
    par_ba = ifelse(x1paredu >= 4, 1, 0),

    # female: binary numeric (1 = female)
    female = ifelse(x1sex == 2, 1, 0),

    # frpl: binary numeric
    frpl = ifelse(x1poverty185 == 0, "0 no", "1 yes"),

    # pov_level: ordinal numeric (1 = deepest poverty → 4 = no poverty)
    pov_level = ordered(
      case_when(
        x1poverty185 == 0  ~ 4,
        x1poverty185 == 1 & x1poverty130 == 0 ~ 3,
        x1poverty130 == 1 & x1poverty == 0 ~ 2,
        x1poverty == 1 ~ 1
      ),
      levels = c(1, 2, 3, 4)
    ),

    # first_lang: nominal categorical
    first_lang = factor(
      case_when(
        x1duallang == 1 ~ "English only",
        x1duallang == 2 ~ "Non-English only",
        x1duallang == 3 ~ "English and another equally"
      )
    ),

    # esl_program: binary numeric
    esl_program = case_when(
      p1ellever == 1 ~ "1 yes",
      p1ellever == 2 ~ "0 no"
    ),

    # stud_race: nominal categorical
    stud_race = factor(
      case_when(
        x1race %in% c(1, 7) ~ "Indigenous",
        x1race == 2        ~ "Asian",
        x1race == 3        ~ "Black",
        x1race %in% c(4,5) ~ "Latinx",
        x1race == 6        ~ "Multiple",
        x1race == 8        ~ "White"
      )
    ),

    # par_race: nominal categorical
    par_race = factor(
      case_when(
        x1par1race == 2        ~ "Asian",
        x1par1race == 3        ~ "Black",
        x1par1race %in% c(4,5) ~ "Latinx",
        x1par1race == 6        ~ "Multiple",
        x1par1race == 8        ~ "White",
        x1par1race == 9        ~ "Indigenous"
      )
    )
  )

# ---------------------------
# Parent involvement
# ---------------------------
hsls <- hsls %>%
  mutate(
    # par_mtg: binary numeric
    par_mtg = ifelse((p1ptomtg == 1 | p1schmtg == 1), 1, 0),

    # par_sch_inv: count variable (numeric, Poisson-type)
    par_sch_inv = par_mtg +
                  p1schevent +
                  p1volunteer +
                  p1fundraise +
                  p1ptconfer
  )

# ---------------------------
# Parent expectations
# ---------------------------
hsls <- hsls %>%
  mutate(
    able_ba1 = 5 - p1ableba,# reverse code this so that higher numbers are higher ratings of student ability to complete BA
    exp_ba1 = ifelse(p1eduexpect >= 6, 1, 0),
    # par_exp: ordinal numeric (low → high expectations)
    par_exp1 = ordered(
      case_when(
        p1eduexpect == 11            ~ 1,
        p1eduexpect %in% c(1,2,3,4)  ~ 2,
        p1eduexpect %in% c(5,6)      ~ 3,
        p1eduexpect %in% c(7,8)      ~ 4,
        p1eduexpect %in% c(9,10)     ~ 5
      ),
      levels = c(1, 2, 3, 4, 5)
    )
  )

# check poverty groups
##hsls %>% dplyr::select(pov_level, starts_with("x1pov")) %>% gtsummary::tbl_summary(by = pov_level)
```

# Factor Analysis of Parent Data

## Subset Parent Questions for Base Year and followup
```{r}
df_p1_home <- hsls %>% dplyr::select(p1eduexpect, p1eduaspire, able_ba1, p1stemdisc, s1momtalkm, s1dadtalkm, s1momtalks, s1dadtalks)

df_p2_home <- hsls %>% dplyr::select(p2eduexp, p2eduasp, p2stemdisc, p2disccourses)

df_p1_school <- hsls %>% dplyr::select(p1ptomtg, p1ptomtg, p1schevent, p1schmtg, p1volunteer, p1fundraise, p1counselor)

df_p <- hsls %>% dplyr::select(p1eduexpect, p1eduaspire, p1ableba, p1stemdisc, s1momtalkm, s1dadtalkm, s1momtalks, s1dadtalks, p2eduexp, p2eduasp, p2ableba, p2stemdisc, p2disccourses)
```

## load factor analysis and scale related libraries
```{r}
library(psych)
library(EFAtools)
library(lavaan)
```

## Test # of Factors and Suitability for Factor Analysis
```{r}
N_FACTORS(df_p1_home) # 2 factors
N_FACTORS(df_p2_home, criteria = c("PARALLEL", "EKC", "SMT")) # 2 factors, need to be more specific in call since less than 6 variables
N_FACTORS(df_p1_school) # 1 or 2 factors
```

## Exploratory Factor Analysis
### Parent Home BY
```{r}
fit_p1_home <- EFA(df_p1_home, 
    n_factors = 2, # two factors
    rotation = "promax", 
    cor_method = "spearman", # b/c Likert
    method = c("PAF", "ML", "ULS"))
fit_p1_home
# save factor loadings as csv file
fit_p1_home_data <- cbind(fit_p1_home$rot_loadings, fit_p1_home$h2) %>% as.data.frame() 
```

### Parent Home FY
```{r}
fit_p2_home <- EFA(df_p2_home, 
    n_factors = 2, # two factors
    rotation = "promax", 
    cor_method = "spearman", # b/c Likert
    method = c("PAF", "ML", "ULS"))
fit_p2_home
# save factor loadings as csv file
fit_p2_home_data <- cbind(fit_p2_home$rot_loadings, fit_p2_home$h2) %>% as.data.frame() 
```

### Parent School BY
```{r}
fit_p1_school <- EFA(df_p1_school, 
    n_factors = 1, # one factor
    rotation = "promax", 
    cor_method = "spearman", # b/c Likert
    method = c("PAF", "ML", "ULS"))
# save factor loadings as csv file
fit_p1_school_data <- cbind(fit_p1_school$rot_loadings, fit_p1_school$h2) %>% as.data.frame() 
fit_p1_school_data
```

## Teacher EFA
### Check Number of Factors
```{r}
# what teachers emphasize in class
mtch_focus <- hsls %>% dplyr::select(m1interest, m1concepts, m1algorithm, m1compskills, m1problem, m1reason, m1ideas, m1prepare, m1logic, m1history, m1explain, m1business, m1compute, m1test)
N_FACTORS(mtch_focus)

# efficacy variables
mtch_efficacy <- hsls %>% dplyr::select(m1family, m1discipline, m1stuachieve, m1parent, m1retain, m1redirect, m1getthru, m1homefx)
N_FACTORS(mtch_efficacy)

```

### Run Teacher Focus EFA
```{r}
fit_mtch_focus <- EFA(mtch_focus, 
    n_factors = 3, # two factors
    rotation = "promax", 
    cor_method = "spearman", # b/c Likert
    method = c("PAF", "ML", "ULS"))
fit_mtch_focus
# save factor loadings as csv file
fit_mtch_focus_data <- cbind(fit_mtch_focus$rot_loadings, fit_mtch_focus$h2) %>% as.data.frame() 
```

### Run Teacher Efficacy EFA
```{r}
fit_mtch_efficacy <- EFA(mtch_efficacy, 
    n_factors = 2, # two factors
    rotation = "promax", 
    cor_method = "spearman", # b/c Likert
    method = c("PAF", "ML", "ULS"))
fit_mtch_efficacy
# save factor loadings as csv file
fit_mtch_efficacy_data <- cbind(fit_mtch_efficacy$rot_loadings, fit_mtch_efficacy$h2) %>% as.data.frame() 
```
# Confirmatory Factor Analysis
```{r}
library(lavaan)

## deficit framing
m1 <- 'deficit =~ m1family +  m1discipline + m1stuachieve + m1parent + m1homefx'
deficit_cfa <- cfa(m1, data = hsls)
summary(deficit_cfa)

# Efficacy of Pedagogy
m2 <- 'efficacy =~ m1stuachieve + m1retain + m1redirect + m1getthru'
efficacy_cfa <- cfa(m2, data = hsls)
summary(efficacy_cfa)

```

# Structural Equation Modelling
```{r}
sem_model <- '
# Factors
deficit =~ m1family +  m1discipline + m1stuachieve + m1parent + m1homefx
par_home =~ p1eduexpect + p1eduaspire + able_ba1
par_discuss =~ s1momtalkm + s1dadtalkm + s1momtalks + s1dadtalks 

# Regressions
deficit ~ par_sch_inv + par_home + adv_math + frpl
x3tcredmat ~ par_sch_inv + par_home + par_discuss + frpl + par_ba + female + math_score + adv_math
s1dadtalkm	~~	s1dadtalks
s1momtalkm	~~	s1momtalks
s1dadtalkm	~~	s1momtalks
s1momtalkm	~~	s1dadtalks
m1family	~~	m1homefx
p1eduexpect	~~	p1eduaspire
'
fit <- sem(sem_model, data = hsls)
summary(fit, fit.measures = TRUE)
modindices(fit, sort = TRUE)
```

# Save Clean Dataset

```{r save}
write_dta(hsls, "hsls_subset.dta")
```




