---
title: "parent_analysis_special_issue"
output: word_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Load libraries and dataset
```{r}
library(tidyverse)
library(gtsummary)
library(jtools)
library(psych)
library(EFAtools)
library(lavaan)
df_subset <- haven::read_dta("hsls_subset_wide.dta") # can use numeric values

```
# Reliability and Validity

## Subset Parent Questions for pa Year
```{r}
psd <- df_subset %>% dplyr::select(s1momtalkm, s1dadtalkm, s1momtalks, s1dadtalks)
phbae <- df_subset %>% dplyr::select(par_exp1, par_asp1, able_ba1)
home <- df_subset %>% dplyr::select(par_exp1, par_asp1, able_ba1, s1momtalkm, s1dadtalkm, s1momtalks, s1dadtalks)
school <- df_subset %>% dplyr::select(p1schevent, p1volunteer, p1fundraise, p1ptconfer, 
             p1ptomtg, p1schmtg)
```

## Reliability
```{r}
psych::reliability(psd)
psych::reliability(phbae)
psych::reliability(home)
psych::reliability(school)

```

# Unimputed

## Prep data
```{r}
df_analysis <- df_subset %>% mutate(
  home = rowMeans(
    dplyr::select(., par_exp1, par_asp1, able_ba1, 
           s1momtalkm, s1dadtalkm, s1momtalks, s1dadtalks)),
  school = rowSums(
    dplyr::select(., p1schevent, p1volunteer, p1fundraise, 
           p1ptconfer, p1ptomtg, p1schmtg)),
    asian_ec = case_when(
      x1race == 2 ~ 1,
      x1race == 8 ~ -1,
      TRUE ~ 0
    ),
    
    black_ec = case_when(
      x1race == 3 ~ 1,
      x1race == 8 ~ -1,
      TRUE ~ 0
    ),
    
    latinx_ec = case_when(
      x1race %in% c(4,5) ~ 1,
      x1race == 8 ~ -1,
      TRUE ~ 0
    ),
    
    
    multiracial_ec = case_when(
      x1race == 6 ~ 1,
      x1race == 8 ~ -1,
      TRUE ~ 0
    ),
    
    another_ec = case_when(
      x1race %in% c(1,7) ~ 1,
      x1race == 8 ~ -1,
      TRUE ~ 0
    )
  )
df_analysis <- df_analysis %>% haven::as_factor() #make it factors
```

## Ordinal Regression
```{r}
# load library for ordinal regression
library(ordinal)

# set Asian as racial reference
df_analysis$x1race <- fct_relevel(df_analysis$x1race, "Asian, non-Hispanic")

# run ordinal regression
olr <- clm(data = df_analysis,
 ordered(highest_math) ~
    psd_scale + phbae_scale + I(scale(par_sch_inv)) + # parent factors
   asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec +  
   x1ses + female + # ses/demographics
    as.numeric(math_level1) + math_score + # prior achievement
    # x1mthid + x1mtheff + x1mthint + x1mthuti + # affect
    iss + urban + public) # school level

# get table
olr %>% tbl_regression(exp = TRUE)   %>% 
  add_glance_source_note() %>% 
  add_significance_stars(hide_p = TRUE, hide_se = FALSE, hide_ci = FALSE) %>% 
  as_flex_table()
```



## Logistic Regression
```{r}
glm <- glm(
           advanced_mathematics ~ 
             psd_scale + phbae_scale + I(scale(par_sch_inv)) + # parent factors
             asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec +
             x1ses + female + # ses/demographics
             early_algebra + as.numeric(math_level1) + math_score + # prior achievement
             #x1mthid + x1mtheff + x1mthint + x1mthuti + # affect
             iss + urban + public, # school level
           data = df_analysis,
           family = binomial)

export_summs(glm, exp = TRUE)

glm %>% tbl_regression(exp = TRUE) %>% 
  add_glance_source_note() %>% 
  add_significance_stars(hide_p = TRUE, hide_se = FALSE, hide_ci = FALSE) %>% 
  as_flex_table()

```
# Weighted Ordinal Regression

## Weight Model
```{r}
library(survey)

df_brr <- svrepdesign(
  weights = ~w3hstrans, 
  repweights = "w3hstrans[001-200]+",
  type = "BRR", 
  data = df_analysis,
  combined.weights = TRUE
)


```
## Ordinal Regression
```{r}

# run ordinal regression
svy_olr <- svyolr(design = df_brr,
 ordered(highest_math) ~
    psd_scale + phbae_scale + I(scale(par_sch_inv)) + # parent factors
    asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec + 
   x1ses + female + # ses/demographics
    early_algebra + as.numeric(math_level1) + math_score + # prior achievement
    # x1mthid + x1mtheff + x1mthint + x1mthuti + # affect
    iss + urban + public) # school level

# get table
svy_olr %>% tbl_regression(exp = TRUE)   %>% 
  add_glance_source_note() %>% add_global_p() %>% 
  add_significance_stars(hide_p = TRUE, hide_se = FALSE, hide_ci = FALSE) %>% 
  as_flex_table()


```



## Logistic Regression
```{r}
svy_glm <- svyglm(design = df_brr,
           advanced_mathematics ~ 
             psd_scale + phbae_scale + I(par_sch_inv) + # parent factors
             asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec + 
             x1ses + female + # ses/demographics
             early_algebra + as.numeric(math_level1) + math_score + # prior achievement
             #x1mthid + x1mtheff + x1mthint + x1mthuti + # affect
             iss + urban + public, # school level
           family = quasibinomial)

# export_summs(svy_glm, exp = TRUE)

svy_glm %>% tbl_regression(exp = TRUE) %>% 
  add_glance_source_note() %>% 
  add_significance_stars(hide_p = TRUE, hide_se = FALSE, hide_ci = FALSE) %>% 
  as_flex_table()

```

# Imputed Data
```{r}
library(mitools)
library(mice)
load("df_imp.rdata")
```

## Unweighted
### Prep
```{r}

# standardize scales
df_imp$data$par_sch_inv_z <- scale(df_imp$data$par_sch_inv)
df_imp$data$phae_z <- scale(df_imp$data$phae)
df_imp$data$psd_z <- scale(df_imp$data$psd)

# make math level numeric
df_imp$data$math_level1_num <- as.numeric(df_imp$data$math_level1)


df_imp$data$asian_ec <- ifelse(df_imp$data$x1race == 2, 1,
                        ifelse(df_imp$data$x1race == 8, -1, 0))

df_imp$data$black_ec <- ifelse(df_imp$data$x1race == 3, 1,
                        ifelse(df_imp$data$x1race == 8, -1, 0))

df_imp$data$latinx_ec <- ifelse(df_imp$data$x1race %in% c(4, 5), 1,
                         ifelse(df_imp$data$x1race == 8, -1, 0))

df_imp$data$multiracial_ec <- ifelse(df_imp$data$x1race == 6, 1,
                              ifelse(df_imp$data$x1race == 8, -1, 0))

df_imp$data$another_ec <- ifelse(df_imp$data$x1race %in% c(1, 7), 1,
                          ifelse(df_imp$data$x1race == 8, -1, 0))


```

### Ordinal Regression
```{r}
# load library for ordinal regression
library(ordinal)

# run ordinal regression
olr_imp <- with(df_imp, clm(
 ordered(highest_math) ~
    psd_scale + phbae_scale + I(scale(par_sch_inv)) + # parent factors
    asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec + 
   x1ses + female + # ses/demographics
    as.numeric(math_level1) + math_score + # prior achievement
    # x1mthid + x1mtheff + x1mthint + x1mthuti + # affect
    iss + urban + public)) # school level

# get table
olr_imp %>% tbl_regression(exp = TRUE)   %>% 
  add_glance_source_note() %>% 
  add_significance_stars(hide_p = TRUE, hide_se = FALSE, hide_ci = FALSE) %>% 
  as_flex_table()

pool(olr_imp)$pooled
```


## Weighted
### BRR Weight Data Imputed
```{r}
# Creates a list of completed dataset for survey weights
imp_list <- imputationList(
  lapply(1:df_imp$m, function(i) complete(df_imp, i))
)

# creates BRR survey design
df_brr_imp <- svrepdesign(
  weights = ~w3hstrans, 
  repweights = "w3hstrans[001-200]+",
  type = "BRR", 
  data = imp_list,
  combined.weights = TRUE
)

# function to exponentiate results and create dataframe with or, se, ll, ul, & p-value
make_mi_table <- function(model) {
  as.data.frame(summary(MIcombine(model))) %>%  # make dataframe of pooled model
  tibble::rownames_to_column("variable") %>% # make variable column
  mutate(
    or = exp(results), # exponentiate estimate for ORs
    ll = exp(`(lower`), # exponentiate ll
    ul = exp(`upper)`), # exponentiate ul
    z = results / se, 
    p.value = 2 * (1 - pnorm(abs(z)))) # calculate p-value with Wald Test
}
```

### Ordinal Regression
```{r}
library(flextable)
start_time <- Sys.time()
# run ordinal regression
olr_imp_svy <- with(df_brr_imp, svyolr(
 ordered(highest_math) ~
    psd_scale + phbae_scale + I(scale(par_sch_inv)) + # parent factors
        asian_ec + black_ec + another_ec + latinx_ec + multiracial_ec + 
   x1ses + female + # ses/demographics
    as.numeric(math_level1) + math_score + # prior achievement
    iss + urban + public)) # school level
end_time <- Sys.time()
end_time - start_time
# get table
olr_imp_svy_tbl <- make_mi_table(olr_imp_svy)
write_csv(olr_imp_svy_tbl, "brr_results.csv")
olr_imp_svy_tbl %>% dplyr::select(variable, or, se, ll, ul, p.value) %>% 
  mutate(
    SE = round(se, 2),
    LL = round(ll, 2),
    UL = round(ul, 2),
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""),
    or_round = round(or, 2),
    OR = paste0(formatC(or_round, format = "f", digits = 2), stars)
    ) %>% dplyr::select(variable, OR, SE, LL, UL) %>% 
  flextable() 

```
